A survey of rigour and reporting standards for gene set enrichment tests

### 2021-08-11 example.Rmd
added stuff

---
title: "PMC Survey Data Analysis"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

@@

library("getDEE2") 
library("DESeq2")
library("clusterProfiler")
library("mitch")
library("kableExtra")
library("eulerr")

@@

```{r,getdata}

mdat<-getDEE2Metadata("hsapiens")

#SRP128998
#samplesheet<-mdat[which(mdat$GEO_series=="GSE109140"),]
samplesheet <- mdat[grep("SRP128998",mdat$SRP_accession),]
samplesheet<-samplesheet[order(samplesheet$SRR_accession),]
samplesheet$HG<-as.factor(c(1,1,1,1,1,1,0,0,0,0,0,0))
samplesheet$VPA<-as.factor(c(0,0,0,1,1,1,0,0,0,1,1,1))

# samplesheets
s1 <- subset(samplesheet,VPA==0)

@@

@@

@@

download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
    destfile="ReactomePathways.gmt.zip")
unzip("ReactomePathways.gmt.zip")
genesets<-gmt_import("ReactomePathways.gmt")

@@

```{r,mitch1}
library(kableExtra)

m <- mitch_import(LGvHG,DEtype = "DEseq2", geneTable = gt)
head(m)
mres <- mitch_calc(m,genesets = genesets)
head(mres$enrichment_result,50)
m_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)[,1]
m_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)[,1]
length(m_up)
length(m_dn)

@@

head(LGvHG)

@@

c_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg, TERM2GENE = genesets2)
c_up <- head(c_up,1000)
head(c_up,20)
c_up <- rownames(subset(c_up, p.adjust < 0.05))

c_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg, TERM2GENE = genesets2)
c_dn <- head(c_dn,1000)
head(c_up,20)
c_dn <- rownames(subset(c_dn, p.adjust < 0.05))

@@

@@

f_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg, TERM2GENE = genesets2)
f_up <- head(f_up,1000)
head(f_up,20)
f_up <- rownames(subset(f_up, p.adjust < 0.05))

f_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg, TERM2GENE = genesets2)
f_dn <- head(f_dn,1000)
head(f_up,20)
f_dn <- rownames(subset(f_dn, p.adjust < 0.05))

```
## Venn diagram comparison

@@

@@

plot(euler(v1),quantities = TRUE)

@@

plot(euler(v2),quantities = TRUE)

v3 <- list("FCS up"=m_up, "FCS dn"=m_dn,
           "ORA up"=c_up,"ORA dn"=c_dn, 
           "ORA* up"=f_up,"ORA* dn"=f_dn )

plot(euler(v3),quantities = TRUE)

>>>>

---
title: "Example gene set analysis"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

@@

suppressPackageStartupMessages({
library("getDEE2") 
library("DESeq2")
library("clusterProfiler")
library("mitch")
library("kableExtra")
library("eulerr")
})

@@

```{r,getdata,fig.width=7,fig.height=7}

mdat<-getDEE2Metadata("hsapiens")
samplesheet <- mdat[grep("SRP128998",mdat$SRP_accession),]
samplesheet<-samplesheet[order(samplesheet$SRR_accession),]
samplesheet$HG<-as.factor(c(1,1,1,1,1,1,0,0,0,0,0,0))
samplesheet$VPA<-as.factor(c(0,0,0,1,1,1,0,0,0,1,1,1))
s1 <- subset(samplesheet,VPA==0)

@@

# MDS plot
plot(cmdscale(dist(t(x1))), xlab="Coordinate 1", ylab="Coordinate 2", pch=19, col=s1$HG, main="MDS")

@@

```{r,deplots,fig.width=7,fig.height=7}

maplot <- function(de,contrast_name) {
  sig <-subset(de, padj < 0.05 )
  up <-rownames(subset(de, padj < 0.05 & log2FoldChange > 0))
  dn <-rownames(subset(de, padj < 0.05 & log2FoldChange < 0))
  GENESUP <- length(up)
  GENESDN <- length(dn)
  DET=nrow(de)
  SUBHEADER = paste(GENESUP, "up, ", GENESDN, "down", DET, "detected")
  ns <-subset(de, padj > 0.05 )
  plot(log2(de$baseMean),de$log2FoldChange, 
       xlab="log2 basemean", ylab="log2 foldchange",
       pch=19, cex=0.5, col="dark gray",
       main=contrast_name, cex.main=0.7)
  points(log2(sig$baseMean),sig$log2FoldChange,
         pch=19, cex=0.5, col="red")
  mtext(SUBHEADER,cex = 0.7)
}

make_volcano <- function(de,name) {
    sig <- subset(de,padj<0.05)
    N_SIG=nrow(sig)
    N_UP=nrow(subset(sig,log2FoldChange>0))
    N_DN=nrow(subset(sig,log2FoldChange<0))
    DET=nrow(de)
    HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn", DET, "detected")
    plot(de$log2FoldChange,-log10(de$padj),cex=0.5,pch=19,col="darkgray",
        main=name, xlab="log2 FC", ylab="-log10 pval", xlim=c(-6,6))
    mtext(HEADER)
    grid()
    points(sig$log2FoldChange,-log10(sig$padj),cex=0.5,pch=19,col="red")
}

maplot(LGvHG,"LG v HG")
make_volcano(LGvHG,"LG v HG")

@@

if (! file.exists("ReactomePathways.gmt")) {
  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
    destfile="ReactomePathways.gmt.zip")
  unzip("ReactomePathways.gmt.zip")
}
genesets<-gmt_import("ReactomePathways.gmt")

@@

```{r,mitch1}

m <- mitch_import(LGvHG,DEtype = "DEseq2", geneTable = gt)
head(m)
mres <- mitch_calc(m,genesets = genesets)
head(mres$enrichment_result) 
m_up <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)[,1]
m_dn <- subset(mres$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)[,1]
length(m_up)
length(m_dn)

@@

@@

c_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg, TERM2GENE = genesets2)
c_up <- head(c_up,1000)
head(c_up)
c_up <- rownames(subset(c_up, p.adjust < 0.05))

c_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg, TERM2GENE = genesets2)
c_dn <- head(c_dn,1000)
head(c_dn)
c_dn <- rownames(subset(c_dn, p.adjust < 0.05))

## ORA with clusterprofiler combining up and down

```{r,cp2}

LGvHG_de <- rownames(subset(LGvHG,padj<0.05))
LGvHG_de <- unique(gt[which(rownames(gt) %in% LGvHG_de),1])
head(LGvHG_de)

d_de <- enricher(gene = LGvHG_de, universe = LGvHG_bg, TERM2GENE = genesets2)
d_de <- head(d_de,1000)
d_de <- rownames(subset(d_de, p.adjust < 0.05))
head(d_de)

@@

```{r,cp3}

@@

f_up <- enricher(gene = LGvHG_up, universe = LGvHG_bg, TERM2GENE = genesets2)
f_up <- head(f_up,1000)
head(f_up)
f_up <- rownames(subset(f_up, p.adjust < 0.05))

f_dn <- enricher(gene = LGvHG_dn, universe = LGvHG_bg, TERM2GENE = genesets2)
f_dn <- head(f_dn,1000)
head(f_dn)
f_dn <- rownames(subset(f_dn, p.adjust < 0.05))

## ORA combined with clusterprofiler with whole genome background list

```{r,cp4}

e_de <- enricher(gene = LGvHG_de, universe = LGvHG_bg, TERM2GENE = genesets2)
e_de <- head(e_de,1000)
head(e_de)
e_de <- rownames(subset(e_de, p.adjust < 0.05))

## Venn diagram comparison

@@

par(cex.main=0.5)

par(mar=c(2,2,2,2))

v0 <- list("ORA up"=c_up,"ORA dn"=c_dn,
           "ORA comb" = d_de)

plot(euler(v0),quantities = TRUE,main="effect of combining up and down regulated genes")

@@

plot(euler(v1),quantities = TRUE, main="FCS compared to ORA")

@@

plot(euler(v2),quantities = TRUE, main="Effect of inappropriate background* (whole genome)")

vx <- list("ORA up"=c_up,"ORA dn"=c_dn,
           "ORA comb" = d_de, "ORA* comb" = e_de)

plot(euler(vx),quantities = TRUE, main="combining up and down genes and whole genome bg*")

v3 <- list("ORA up"=c_up,"ORA dn"=c_dn, 
           "ORA* up"=f_up,"ORA* dn"=f_dn ,
           "FCS up"=m_up, "FCS dn"=m_dn)

plot(euler(v3),quantities = TRUE, main="all sets")

## Jaccard calculation

```{r,jacc1}

# ORA vs ORA combined
dc <- length(intersect(d_de, c(c_up,c_dn))) / length(union(d_de, c(c_up,c_dn)))
dc

# FCS vs ORA
cm <- length(intersect(c(c_up,c_dn), c(m_up,m_dn))) / length(union(c(c_up,c_dn), c(m_up,m_dn)))
cm

m_up <- gsub("^","up ",m_up)
m_dn <- gsub("^","dn ",m_dn)
m_de <- union(m_up,m_dn)

c_up <- gsub("^","up ",c_up)
c_dn <- gsub("^","dn ",c_dn)
c_de <- union(c_up,c_dn)

f_up <- gsub("^","up ",f_up)
f_dn <- gsub("^","dn ",f_dn)
f_de <- union(f_up,f_dn)

# ORA vs ORA*
cf <- length(intersect(c_de, f_de )) / length(union(c_de, f_de))
cf

# FCS vs ORA*
mf <- length(intersect(m_de, f_de )) / length(union(m_de, f_de))
mf

dat <- c("ORA vs ORA comb"=dc,"FCS vs ORA"=cm,"ORA vs ORA*"=cf,"FCS vs ORA*"=mf)

dat

barplot(dat,ylab="jaccard metric")

### 2021-08-11 manuscript.Rmd
added stuff

>>>>

As we screened these articles, we observed other types of details lacking.
For example GSEA has different options that impact the results including the ranking metric, gene weighting method and the permutation type (on samples or genes), which were not regularly reported in articles.
For ORA, in some instances it was unclear whether up- and down-regulated gene lists were combined before analysis.
For differential expression analysis, combining up- and down-regulated gene lists for ORA is considered poor practice (Hong et al, 2014).
Our own tests indicate that combining up- and down-regulated gene lists prior to ORA dramatically reduces the statistical strength when the proportion of differentially expressed genes is high.
